---
title: "05_analysis"
format: html
editor: visual
---

# Microbial distribution across BMI classes

```{r}
library(tidyverse)
library(vegan)
```

```{r}
# verify that the dataset can be loaded successfully
data_path <- "../data/06_dat_augmented.csv"
dataset <- read_csv(data_path,
                    show_col_types = FALSE)
```

We want to investifate if there is a significant difference in term of microbial communities in the different BMI groups. To do so we first visualise with PCoA if the microbial composition form clusters when plotted. Secondly we perform a PERMANOVA test to quantify the difference between the BMI groups, testing the following:

Null hypothesis : There is no difference in microbial community across composition across the BMI classes

Alternative hypothesis: There is a difference in microbial community composition across BMI classes.

```{r}
#splitting dataset into metadata and bacteria data 
metadata <- dataset |> 
  select(-starts_with("s_")) 

bacteria_data <- dataset |> 
  select(starts_with("s_")) 
```

```{r}
#  Calculate the Bray-Curtis distances
bc_distance <- vegdist(bacteria_data,
                   method = "bray")
```

```{r}
# applying (PCoA) to visualize clusters of bacteria.
pcoa_result <- cmdscale(bc_distance, k = 2)  
```

```{r}
# PCoA with BMI to visualize clusters across BMI
pcoa_df <- data.frame(pcoa_result,
                      Group = metadata$BMI_class) |> 
  rename(PCoA1 = X1,
         PCoA2 = X2)
```

```{r}
 ggplot(pcoa_df,
       aes(x = PCoA1,
           y = PCoA2,
           color = Group)) +
    geom_point(size = 3,
             alpha = 0.5) +  
    theme_minimal() +  
    labs(title = "PCoA Plot of Beta Diversity", 
       x = "PCoA1", 
       y = "PCoA2",
       color = "BMI Category")
```

```{r}
 # PERMANOVA to  see whether the diversity of microbial communities differ across bmi
anova_result <- adonis2(bc_distance ~ BMI_class,
                         data = metadata)
print(anova_result)
```

## Interpretation of the results

From the PCoA plot we can see that there aren't clusters for each BMI category and some samples overlap with samples from different categories such as points from normal weight overlap with obese and overweight.

The results of the PERMANOVA agree with what is seen in the PCoA as the P.value is 0.219, which indicates that the differences in microbial composition across BMI classes is not significant. Also R2=0.0238 means that bmi contributes to only 2.38% of the variance in the diversity of microbial community across the bmi classes.
