---
title: "05_analysis"
author: "Andreis Marco"
format: html
editor: visual
---

# Load the dataset and required packages

```{r}
library(tidyverse)
library(vegan)
```

```{r}
# verify that the dataset can be loaded successfully
data_path <- "../data/06_dat_augmented.csv"
dataset <- read_csv(data_path,
                    show_col_types = FALSE)
```

### Metadata analysis

What does impaired glucose tolerance mean? It means that blood sugar levels are elevated but not enough to warrant a diagnosis of diabetes

```{r}
# Let's look at the proportions of our catagorical values
# Our data is well proportionally split between the three distincs groups of "disease"
dataset |> 
  count(glucose_tollerance,
        name = "Count") |> 
  arrange(desc(Count))
```

### Correlation analysis

Calculating the correlation matrix on a subset of our data will tell us about linear relationships between data variables.

Lets start with a subset of our data, containing metadata and a few high-level markers (e.g. LDL and HDL)

```{r}
# Filter for diabetics and impaired
filtered_data <- dataset |> 
  filter(glucose_tollerance %in% c("t2d", "impaired"))

# Choose the abundance data from the filtered data
df_selected <- filtered_data |> 
  select(starts_with("s_"))

# Calculate the mean of each column
mean_values <- df_selected |> 
  summarise(across(everything(),
                   mean))

# Convert to long format and get the top 5 columns with the highest mean values
# top_5_species <- mean_values |> 
#   pivot_longer(everything(),
#                names_to = "species",
#                values_to = "mean_value") |> 
#   arrange(desc(mean_value)) |> 
#   slice_head(n = 5) |> 
#   pull(species)
# 
# top_5_species <- dataset |> 
#   select(all_of(top_5_species))
```

```{r}
corr_data <- dataset |> 
  select(-starts_with("s_")) |> 
  select(where(~is.numeric(.x))) 

# Normalize  columns
corr_normalized <- corr_data |> 
  mutate(across(everything(),
                ~ (. - mean(.)) / sd(.)))

# Compute the covariance matrix
cov_matrix <- (t(as.matrix(corr_normalized)) %*% as.matrix(corr_normalized)) / 
  (nrow(corr_normalized) - 1)

# Convert the covariance matrix to a correlation matrix
cor_matrix <- cov2cor(cov_matrix)

# Pivot the correlation matrix to long format
melted_cor_matrix <- as.data.frame(cor_matrix) |> 
  rownames_to_column(var = "Var1") |> 
  pivot_longer(-Var1, names_to = "Var2"
               , values_to = "value")

# Plot values in colour squares to visualize correlation
ggplot(data = melted_cor_matrix,
       aes(x = Var1,
           y = Var2,
           fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value
                              , 2)),
            color = "black",
            size = 2) + 
  scale_fill_gradient2(low = "blue",
                       high = "red",
                       mid = "white", 
                       midpoint = 0,
                       limit = c(-1, 1),
                       space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle =90,
                                   vjust = 1, 
                                   size = 10,
                                   hjust = 1),
  axis.title.x = element_blank(),
  axis.title.y = element_blank())+
  coord_fixed()
```

We see a slight inverse relationship between HDL levels and BMI. This might not come as the biggest surprise. \[doi: 10.1586/erc.11.182\]

## Assessing relationships between species abundance and metadata features

Let's analyze the diversity between class variables (T2d, healthy and glucose impaired)

We do this by calculating a distance matrix based on the microbial abundance data, which then can be used to visualize the differences in composition in a reduced dimensionality.

```{r}
# Pull the class labels as a vector 
group_labels <- dataset |> 
  pull(glucose_tollerance)

# Select abundance data from data set to calculate distsances 
abundance_data <- dataset |>
  select(matches("^s_"))

# Calculate Beta diversity based on Bray-Curtis distance
bc_dist <- vegdist((abundance_data),
                   method = "bray")

# Perform PCoA (Principal Coordinates Analysis)
pcoa_result <- cmdscale(as.dist(bc_dist),
                        k = 2)

# Combine PCoA results with metadata (group labels)
pcoa_df <- data.frame(pcoa_result,
                      Group = group_labels) |> 
  rename(PCoA1 = X1,
         PCoA2 = X2)
```

```{r}
# PCoA plot
ggplot(pcoa_df,
       aes(x = PCoA1,
           y = PCoA2,
           color = Group)) +
  geom_point(size = 4) +
  theme_minimal() +
  labs(title = "PCoA of Bray-Curtis Dissimilarity",
       x = "PCoA1", 
       y = "PCoA2") +
  scale_color_manual(values = c("red", "blue", "green"))
```

The plot does not show any indication of separation between groups.
