---
title: "06_augment.qmd"
format: html
editor: visual
---

### Load the dataset

```{r}
<<<<<<< HEAD
suppressMessages(library("tidyverse"))
=======
library("tidyverse")
>>>>>>> 4d738e43b3aca70a894ae7d7ee9496308a8fa8e6
library("here")
```

```{r}
# verify that the dataset can be loaded successfully
data_path <- here("data/05_dat_tidy.csv")
<<<<<<< HEAD
dataset <- read_csv(data_path, show_col_types = FALSE)
=======
dataset <- read_csv(data_path,
                    show_col_types = FALSE)
>>>>>>> 4d738e43b3aca70a894ae7d7ee9496308a8fa8e6
```

### Augment

Here we will create a couple of new variables that will come in handy during visualization and analysis.

<<<<<<< HEAD
=======
Create a variable which groups together samples from patients having type 2 diabetes or being in the pre-diabetes state.

```{r}
dataset <- dataset |> 
  mutate(is_diseased = ifelse(glucose_tolerance == "t2d" |
                                glucose_tolerance == "impaired",
                              "yes", "no")
  ) |>
  relocate(is_diseased,
           .after = glucose_tolerance)
```

>>>>>>> 4d738e43b3aca70a894ae7d7ee9496308a8fa8e6
Create a categorical classification based on the BMI.

```{r}
dataset <- dataset |> 
  mutate(BMI_class = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Normal weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
<<<<<<< HEAD
    bmi >= 30 ~ "Obese",
  )) |> 
  relocate(BMI_class, .after = bmi)
```

Create a variable to hold the ratio between high-density and low-density lipoproteins. This ratio is studied and used as a predictor for the insurgence of many cardiovascular diseases.
=======
    bmi >= 30 ~ "Obese"
  ) |> 
  factor(levels = c("Underweight", "Normal weight", "Overweight", "Obese"))) |>
  relocate(BMI_class,
           .after = bmi)
```

Create a variable to hold the ratio between high-density and low-density lipoproteins. This ratio is studied and used as a predictor for the insurgence of many cardiovascular diseases, the desired ratio is \>1 as hdl are the good fats while ldl are the bad ones.
>>>>>>> 4d738e43b3aca70a894ae7d7ee9496308a8fa8e6

```{r}
dataset <- dataset |> 
  mutate(hdl_to_ldl_ratio = hdl / ldl) |> 
<<<<<<< HEAD
  relocate(hdl_to_ldl_ratio, .after = ldl)
=======
  relocate(hdl_to_ldl_ratio,
           .after = ldl)
>>>>>>> 4d738e43b3aca70a894ae7d7ee9496308a8fa8e6
```

### Save dataset

```{r}
target_dir <- "../data"
<<<<<<< HEAD

# Create the data folder if it does not exist
if (!file.exists(target_dir)) {
  # Create the folder if it doesn't exist
  dir.create(target_dir, recursive = TRUE)
  }

# save the dataset in the data fodler
write_csv(dataset, str_c(target_dir, "/06_dat_augmented.csv"))
=======
write_csv(dataset, str_c(target_dir,
                         "/06_dat_augmented.csv"))
>>>>>>> 4d738e43b3aca70a894ae7d7ee9496308a8fa8e6
```
